Narrative

### 2023-05-25 manuscript/manuscript.qmd
manuscript: added figure captions

Figure 1 show the absolute and relative bias in heterogeneity estimates for different levels of heterogeneity (measured in $\tau$) and meta-analytic models. Each column of Figure 1 corresponds to one nominal level of heterogeneity ($\tau$ $\epsilon$ {0, 0.1, 0.15, 0.2}) and each row to a different analysis (HV with Fisher $z$, HV with Pearson's $r$, and HS). The black unbroken lines indicate actual heterogeneity, which may differ from nominal heterogeneity because of truncation or in the case of Fisher $z$ because the comparable heterogeneity levels were slightly different (see methods). The different dashed lines in Figure 1 shows the $\hat{\tau}$ estimates for four different levels of average reliability across studies, and the x-axis shows the average (superpopulation) effect size. As can be seen, the bias is marginally worse for Fisher $z$ than the other two methods (although Fisher $z$ is on a different scale) but the results are approximately the same for all three methods. As such, we will focus in the remainder of the results on Pearson's $r$ (row two).

@@

_Figure 2._ EXPLANATION. Code to reproduce figure: LINK.

As expected, for zero heterogeneity (leftmost column Figure 1) we see an overestimate of heterogeneity that increases with effect size. Although Pearson's $r$ has a small positive bias due to the dependance between effect size and sampling variance (supplement XX), more generally there are two sources of positive bias in the absence of heterogeneity. First, and the only one that applies when $\mu = 0$, there is positive bias due to the truncation of negative heterogeneity estimates (Viechtbauer, 2005). Second, there is bias due to the variance in reliability. For example, for $mu = 0.2$ (the closest value to the median correlation reported by Schäfer & Schwarz) and $\bar{R} = 0.8$ the bias for Pearson's $r$ compared to true zero heterogeneity is `r in_text$bias_tau_zero$mu_0.2`. For $\mu = 0.4$ (the upper median reported by Schäfer & Schwarz) and $\bar{R} = 0.8$ the bias is `r in_text$bias_tau_zero$mu_0.4`. For the highest average reliability (0.9) a somewhat less inclined slope can be seen. This is because the standard deviation for this reliability level is severely truncated when the reliability is bounded at 1. The increase in positive bias with effect size can also be seen in the presence of heterogeneity (all other columns). However, it is superseded by the negative bias.

For all heterogeneity levels in Figure 1 above zero we see an overall underestimate of heterogeneity due to unreliability in primary studies. This negative bias can be relatively severe, is worse with lower effect size and lower average reliability, and worsens in an absolute sense but improves in a relative sense as heterogeneity increases. For example, for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.1$, the bias is `r in_text$bias_increasing_tau$tau_0.1` for Pearson's $r$, for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.15$ the bias is `r in_text$bias_increasing_tau$tau_0.15`, and for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.2$ the bias is `r in_text$bias_increasing_tau$tau_0.2`. For an effect size of $\mu = 0.2$ true heterogeneity between 0.1 - 0.2 is underestimated by between `r round(100*in_text$underestimate_mu02$underestimate_min$underestimate, 0)`% ($\tau = 0.2$, $\bar{R} = 0.9$) and `r round(100*in_text$underestimate_mu02$underestimate_max$underestimate, 0)`% ($\tau = 0.1$, $\bar{R} = 0.6$).

To explore further under what conditions the positive and negative bias in heterogeneity estimates would reach an equilibrium, we added an analysis for several conditions with $\tau < 0.1$. We used the following values {0.02, 0.04, 0.06, 0.08}. Figure 2 presents the results of this analysis for Pearson's $r$. Because these lower heterogeneity levels can be expected to be affected by truncation to a larger extent (which results in a positive bias), we present results for four different numbers of studies in the meta-analysis (rows Figure 2, lower $k$ is associated with larger positive bias). The standard deviation in reliabilities is again 0.15. Figure 2 shows that even when only little heterogeneity is present, it will generally be underestimated. If $\mu < 0.3$ then imperfect reliability leads to a negative bias in heterogeneity estimates for all $\tau < 0.06$ in Pearson's $r$. Even for lower levels of heterogeneity, heterogeneity will generally be positively biased only when average effect size is large or the number of meta-analyzed studies is small.

@@

_Figure 3._ EXPLANATION. Code to reproduce figure: LINK.

@@

Similarly, if researchers find zero heterogeneity in their observed estimates, true heterogeneity is most likely to be zero or small. Larger true heterogeneity (e.g., $\tau > 0.08$) has a higher expected observed heterogeneity value even when average reliability is as low as 0.6, and as such is expected to result in zero estimates less often than zero or smaller true heterogeneity for fixed conditions. For example, for the conditions reported in the main article (i.e., K = 20, N = 150, $\sigma_R$ = 0.15), an assuming a typical (reported) average reliability of 0.8 and a zero average correlation, then an observed zero heterogeneity estimate is XX times more likely to correspond to true zero heterogeneity than $\tau = 0.1$. That said, an observed zero heterogeneity estimate is still possible to observe under larger true heterogeneity levels as variability in heterogeneity estimates is often large in practice (Ioannidis; ref2). In addition, separating between zero and small heterogeneity is unfeasible considering realistic sample sizes in meta-analyses in psychology (Olsson-Collentine et al., 2020). In our view, the distinction between zero and small heterogeneity in meta-analysis is primarily a hypothetical one, as we consider low heterogeneity and meta-analysis of a large number of studies unlikely to cooccur in practice.

@@

Imperfect measurement reliability in primary studies generally leads to a severe underestimate of observed meta-analytic heterogeneity. As few meta-analyses in psychology correct for unreliability in primary studies, heterogeneity is likely underestimated in a large proportion of meta-analyses in psychology. Consequently, variability in effects has been ignored, moderators potentially neglected and theoretical development in psychology been misguided by meta-analyses. Sophisticated methods (MASEM) for correcting measurement unreliability are only applicable in exceptional cases and corrections based on classical test theory come with caveats and assumptions. Accurate estimation of meta-analytic heterogeneity is difficult and will remain so unless measurement concerns (REF) are taken seriously in primary research. The good news is that study designs (multi-lab replication studies) which can apply more sophisticated versions of MASEM are becoming increasingly common in psychology, and that large true heterogeneity can on average be distinguished from zero or small heterogeneity even in the absence of corrections to unreliability in primary studies.

>>>>

Figure 2 show the absolute and relative bias in heterogeneity estimates for different levels of heterogeneity (measured in $\tau$) and meta-analytic models. Each column of Figure 2 corresponds to one nominal level of heterogeneity ($\tau$ $\epsilon$ {0, 0.1, 0.15, 0.2}) and each row to a different analysis (HV with Fisher $z$, HV with Pearson's $r$, and HS). The black unbroken lines indicate actual heterogeneity, which may differ from nominal heterogeneity because of truncation or in the case of Fisher $z$ because the comparable heterogeneity levels were slightly different (see methods). The different dashed lines in Figure 2 shows the $\hat{\tau}$ estimates for four different levels of average reliability across studies, and the x-axis shows the average (superpopulation) effect size. As can be seen, the bias is marginally worse for Fisher $z$ than the other two methods (although Fisher $z$ is on a different scale) but the results are approximately the same for all three methods. As such, we will focus in the remainder of the results on Pearson's $r$ (row two).

@@

_Figure 2._ Unreliability in primary studies leads to a net negative bias when heterogeneity is positive.  The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations. Columns indicate the nominal true heterogeneity standard deviation $\tau$. Due to truncation in Pearson's $r$ or translation to Fisher's $z$ these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line correspond to an average reliability in primary studies. Standard deviation in reliability was 0.15. Code to reproduce figure: LINK.

As expected, for zero heterogeneity (leftmost column Figure 2) we see an overestimate of heterogeneity that increases with effect size. Although Pearson's $r$ has a small positive bias due to the dependance between effect size and sampling variance (supplement XX), more generally there are two sources of positive bias in the absence of heterogeneity. First, and the only one that applies when $\mu = 0$, there is positive bias due to the truncation of negative heterogeneity estimates (Viechtbauer, 2005). Second, there is bias due to the variance in reliability. For example, for $mu = 0.2$ (the closest value to the median correlation reported by Schäfer & Schwarz) and $\bar{R} = 0.8$ the bias for Pearson's $r$ compared to true zero heterogeneity is `r in_text$bias_tau_zero$mu_0.2`. For $\mu = 0.4$ (the upper median reported by Schäfer & Schwarz) and $\bar{R} = 0.8$ the bias is `r in_text$bias_tau_zero$mu_0.4`. For the highest average reliability (0.9) a somewhat less inclined slope can be seen. This is because the standard deviation for this reliability level is severely truncated when the reliability is bounded at 1. The increase in positive bias with effect size can also be seen in the presence of heterogeneity (all other columns). However, it is superseded by the negative bias.

For all heterogeneity levels in Figure 2 above zero we see an overall underestimate of heterogeneity due to unreliability in primary studies. This negative bias can be relatively severe, is worse with lower effect size and lower average reliability, and worsens in an absolute sense but improves in a relative sense as heterogeneity increases. For example, for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.1$, the bias is `r in_text$bias_increasing_tau$tau_0.1` for Pearson's $r$, for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.15$ the bias is `r in_text$bias_increasing_tau$tau_0.15`, and for $\bar{R} = 0.8$, $\mu = 0.2$, $\tau = 0.2$ the bias is `r in_text$bias_increasing_tau$tau_0.2`. For an effect size of $\mu = 0.2$ true heterogeneity between 0.1 - 0.2 is underestimated by between `r round(100*in_text$underestimate_mu02$underestimate_min$underestimate, 0)`% ($\tau = 0.2$, $\bar{R} = 0.9$) and `r round(100*in_text$underestimate_mu02$underestimate_max$underestimate, 0)`% ($\tau = 0.1$, $\bar{R} = 0.6$).

To explore further under what conditions the positive and negative bias in heterogeneity estimates would reach an equilibrium, we added an analysis for several conditions with $\tau < 0.1$. We used the following values {0.02, 0.04, 0.06, 0.08}. Figure 3 presents the results of this analysis for Pearson's $r$. Because these lower heterogeneity levels can be expected to be affected by truncation to a larger extent (which results in a positive bias), we present results for four different numbers of studies in the meta-analysis (rows Figure 3, lower $k$ is associated with larger positive bias). The standard deviation in reliabilities is again 0.15. Figure 3 shows that even when only little heterogeneity is present, it will generally be underestimated. If $\mu < 0.3$ then imperfect reliability leads to a negative bias in heterogeneity estimates for all $\tau < 0.06$ in Pearson's $r$. Even for lower levels of heterogeneity, heterogeneity will generally be positively biased only when average effect size is large or the number of meta-analyzed studies is small.

@@

_Figure 3._ Unreliability in primary studies typically leads to a net negative bias even for small degrees of heterogeneity. Effect size is Pearson's $r$ and rows indicate the number of studies in the meta-analysis. The x-axis indicates average effect size and the y-axis estimated heterogeneity in standard deviations. Columns indicate the nominal true heterogeneity standard deviation $\tau$. Due to truncation in Pearson's $r$ or translation to Fisher's $z$ these values may differ from actual true heterogeneity standard deviation (black solid lines). Each dashed line correspond to an average reliability in primary studies. Standard deviation in reliability was 0.15. Code to reproduce figure: LINK.

@@

Similarly, if researchers find zero heterogeneity in their observed estimates, true heterogeneity is most likely to be zero or small. Larger true heterogeneity (e.g., $\tau > 0.08$) has a higher expected observed heterogeneity value even when average reliability is as low as 0.6, and as such is expected to result in zero estimates less often than zero or smaller true heterogeneity for fixed conditions. For example, for the conditions reported in the main article (i.e., K = 20, N = 150, $\sigma_R$ = 0.15), an assuming a typical (reported) average reliability of 0.8 and a zero average correlation, then an observed zero heterogeneity estimate is XX times more likely to correspond to true zero heterogeneity than $\tau = 0.1$. That said, an observed zero heterogeneity estimate is still possible to observe under larger true heterogeneity levels as variability in heterogeneity estimates is often large in practice (Ioannidis; ref2). In addition, separating between zero and small heterogeneity is infeasible considering realistic sample sizes in meta-analyses in psychology (Olsson-Collentine et al., 2020). In our view, the distinction between zero and small heterogeneity in meta-analysis is primarily a hypothetical one, as we consider low heterogeneity and meta-analysis of a large number of studies unlikely to cooccur in practice.

@@

Imperfect measurement reliability in primary studies generally leads to a severe underestimate of observed meta-analytic heterogeneity. As few meta-analyses in psychology correct for unreliability in primary studies, heterogeneity is likely underestimated in a large proportion of meta-analyses in psychology. Consequently, variability in effects has been ignored, moderators potentially neglected and theoretical development in psychology been misguided by meta-analyses. Sophisticated methods (MASEM) for correcting measurement unreliability are only applicable in exceptional cases and corrections based on classical test theory come with caveats and assumptions. Accurate estimation of meta-analytic heterogeneity is difficult and will remain so unless measurement concerns (Flake & Fried, 2020; Flake et al., 2017; Kane et al., 2021) are taken seriously in primary research. The good news is that study designs (multi-lab replication studies) which can apply more sophisticated versions of MASEM are becoming increasingly common in psychology, and that large true heterogeneity can on average be distinguished from zero or small heterogeneity even in the absence of corrections to unreliability in primary studies.
